{Exciton-polariton Josephson junctions at finite temperatures}

### 2017-03-29 writing/phase_transition/phase_transition.tex
Work in progress

\begin{abstract}
We consider finite temperature effects in a non-standard Bose-Hubbard model for an exciton polariton Josephson junction (EPJJ) that is characterised by a complex potential energy landscapes (PEL) consisting of sets of barriers and wells.
We have shown that the transition between  thermal  activation (classical) and tunneling (quantum) regimes exhibits universal features of the first and second order phase transition (PT) depending on the PEL for two polariton condensates.
In the presence of dissipation the relative phase of two condensates exhibits non-equilibrium PT from quantum regime characterized by efficient tunneling of polaritons to the regime of permanent Josephson or Rabi oscillations, where the tunneling is suppressed, respectively.
The obtained results are discussed in details in the framework of the quantum annealing problem with exciton polariton condensates.

@@

In the last decade investigations of exciton polariton Bose-Einstein condensates (BEC) in various type of semiconductor microstructures became and important area of research in photonics and semiconductor physics, see e.g. \cite{Sanvitto, Guillet}.
Microcavity exciton polaritons are bosonic quasiparticles representing admixtures of quantized cavity photons mode and quantum well excitons.
Semiconductor microcavities are promising for various optoelectronic applications where the quantum  matter-field interface plays crucial role.
In particular, it is worth to mention experimental demonstrations of a polariton laser with electrical injection pump \cite{Bhattacharya,Schneider}, amplifiers \cite{Niemietz}, switches \cite{Amo_2010}, transistors \cite{Ballarini}, polariton circuits and optical logic elements \cite{Sturm,Liew}.
Although exciton polariton condensates exhibit the Bose-Einstein statistics above the lasing threshold and are characterized by a macroscopic occupation of the ground state at certain pumping rate that is less than the threshold pumping for convenient lasers, they are not in a true thermodynamic equilibrium state \cite{Byrnes_2014,Sun}. Non-equilibrium features of exciton polariton condensates play an important role in various manifestations of their collective (many body) states such as superfluidity \cite{Carusotto_2013,Amo_2009}, quantized vortices \cite{Lagoudakis_2008,Lagoudakis_2009}, soliton formation \cite{Sich}, Josephson oscillations and macroscopic self-trapping \cite{Abbarchi,Lagoudakis_2010}.

Recently Yongbao Sun et al, in Ref. \cite{Snoke_2017} reported the observation of a quasi-equilibrium low branch exciton polariton condensate in a high Q-factor microcavity possessing a comparatively very long photon lifetime of $135ps$. Notably, exciton polaritons with long lifetime are promising candidates for quantum technologies, cf. \cite{Demirchyan}. However, since exciton polariton BEC's in present experiments are observed at high enough temperatures the quantum properties of such a states should be examined even at full thermal equilibrium.

The problem of distinguishability of statistically classical and quantum regimes for exciton polariton condensates is very important for their possible applications in quantum information technologies, see e.g. \cite{Dominici}.
Fast switching properties (the typical switching time of a few picoseconds), relatively strong nonlinear response and flexibility to external optical and/or electrical pump, spin degrees of freedom made microcavity polaritons potentially very promising for quantum computation and quantum information processing objectives, see, e.g. \cite{Demirchyan,Pagel,Kyriienko,Solnyshkov_2015, Dominici}.

Especially, it is important to emphasize here quantum annealing problem physically based on the searching algorithm for the global minimum of a potential energy landscape (PEL) consisting of a set of barriers and wells, see e.g.  \cite{Santoro, Das}.
In a purely classical  (thermal) regime the {bosonic quasipartcles cross barriers stochastically} at finite temperature with the help of thermal activation process if the thermal energy is large enough. Contrary, in a quantum limit the same system undergoes quantum tunneling through the barrier. Obviously, the shape of the barrier plays an essential role in this case, cf. \cite{Das, Lewenstein}. It has been shown recently in \cite{Lewenstein} that the annealing algorithm relies in general case on combination of thermal annealing and quantum tunneling. The     
collective (bosonic) character of polariton condensation could be used for the acceleration of the physical implementation of the search algorithm \cite{Yan}.

In this paper we consider theoretically the finite temperature effects for exciton polariton condensates formed in a semiconductor microcavity in the presence of the Josephson effect, cf. \cite{Aleiner, Shelykh_2008, Borgh_2010}. 
In particular, we are interesting in generic problem of quantum-classical phase transition (PT) explicitly examined for a two junction model \cite{Chudnovsky_1997}.

In general, this problem is related to the dissipative tunneling problem or tunneling at finite (non-zero) temperatures, cf. \cite{Caldeira, Larkin, Riseborough}.
Traditionally,  studies in this field are limited by consideration of superconductor devices \cite{Ankerhold} where macroscopic quantum tunneling phenomena plays an essential role.
In particular, it is worth to mention here the Schr\"odinger cat state formation \cite{Leggett} and the 	design of quantum gates with superconductor qubits for quantum computing where a macroscopic quantum coherence for the phase is important \cite{Makhlin}.
Recently, the problem of quantum tunneling and quantum-classical PT problem have been examined for other two-level (spin) systems, like atomic BEC's \cite{Zhang}, and small ferromagnetic particles, \cite{Owerre}.

In the paper we consider an \textit{extended model} for exciton polariton Josephson junctions (EPJJs) that relates to so-called non-standard Bose-Hubbard models (see \cite{Dutta} and references therein) when energy of polariton-polariton scattering contributes to the tunneling parameter matrix elements, cf. \cite{Aleiner, Shelykh_2008, Borgh_2010, Solnyshkov_2008, Sarchi}.
In particularly, we pay attention to temperature dependent quantum critical phenomena occurring in the presence of macroscopic tunneling.
We aimed at formulating necessary criteria to obtain quantum phase regimes with trapped exciton polaritons suitable for realisation of quantum annealing algorythms and creation of exciton polariton qubit gates.

This paper is organised as follows.
In Sec. \ref{sec:quantum_phase} we introduce the quantum phase model for extended model of EPJJ of the coupled condensates and discuss various regimes for it at zero temperature and infinite exciton polariton lifetime.
Section \ref{sec:quantum_classical} presents the results obtained for a quantum-classical PT problem and addresses the quantum phase at finite temperatures.
The influence of non-equilibrium effects with microcavity exciton polaritons possessing a finite lifetime is discussed in \ref{sec:non-equilibrium}.      
In Conclusion, we summarize the obtained results.
In the Appendix \ref{sec:model} we explain in details  extended model of EPJJ within the tight-binding approximation.

@@

{\red

We consider a system of two exciton-polariton condensates localised in lateral potential traps in a planar semiconductor microcavity.
The condensates are separated by a potential barrier $U(x)$.
Polaritons are pumped to the condensates from a reservoir of incoherent excitons.
They decay radiatively.
We account for the repulsive intereaction between exciton-polaritons.
The finite temperature of the polariton gas is accounted for in the thermal equilibrium approximation.
We use two-mode representation of the $\hat{\psi}$ - operator that describe the behaviour of the condensates:
%
\begin{equation}
\hat{\psi} = \hat{\psi}_1(t) \Phi_1(x) + \hat{\psi}_2(t) \Phi_2(x).
\end{equation}
%
Here $\Phi_{1,2} = \dfrac{\Phi_+ \pm \Phi_-}{\sqrt{2}}$ where $\Phi_{\pm}(x)$ are symmetric and antisymmetric real wavefuntions obeying the stationary Gross-Pitaevsky equation $\mu_{\pm} \Phi_{\pm} = -\dfrac{\hbar^2}{2m_{pol}} \dfrac{d^2 \Phi_{\pm}}{dx^2} + U(x) \Phi_{\pm} + g \Phi_{\pm}^3$.

%
\begin{figure}[ht]
\center{\includegraphics[width=0.5\linewidth]{pic/potential_picture.eps}}
\caption{ \dots \label{pic:potential_picture}}
\end{figure}
%

We can express the Hamiltonian of such systems in terms of pseudo-spin operators as follows:
%
\begin{equation}
\hat{H} = \alpha \hat{S}_z^2 + \beta \hat{S}_x^2 - B \hat{S}_x
\label{eq:hamiltonian_spin}
\end{equation}
Is characterizes a weakly interacting exciton polariton condensate Josephson junction device.
The parameters $\alpha = 2 \gamma_{+-}$, $\beta = \dfrac{\gamma_{++} + \gamma_{--} - 2 \gamma_{+-}}{2}$ and $B = \mu_- - \mu_+$, $\gamma_{ij} = g \int \Phi_i^2 \Phi_j^2 dx, i,j \in \{+,-\}$ reflect Josephson junction properties, see Appendix \ref{sec:model} for details.
}
Here we propose quantum mechanical approach to the EPJJ  phase recasting pseudo-spin operators in  (\ref{eq:hamiltonian_spin}) through the differential representation defined as
% 
\begin{subequations}
\begin{align}
\hat{S}_x = & s \cos{\phi} - \sin{\phi} \dfrac{d}{d \phi}, \\
\quad \hat{S}_y = & s \sin{\phi} + \cos{\phi} \dfrac{d}{d \phi}, \\
\hat{S}_z = & -i\dfrac{d}{d \phi}
\end{align}
\label{eq:pseudo_spin_diff}
\end{subequations}
%
where  $\phi$ is the phase variable.
The operators determined in Eq. (\ref{eq:pseudo_spin_diff}) in \eqref{eq:hamiltonian_spin}
 obey familiar $SU(2)$ algebra commutation relations $[\hat{S}_i, \hat{S}_j] = i\epsilon_{ijk}\hat{S}_k,  i,j,k=x,y,z$.
After straightforward calculations from (\ref{eq:hamiltonian_spin}) one can obtain
%
\begin{equation}
\begin{array}{lcl}
H & = & -(\alpha - \beta \sin^2 \phi) \dfrac{d^2}{d \phi^2} - (\beta(s-\dfrac{1}{2})\sin{2\phi} - B\sin{\phi}) \dfrac{d}{d \phi} \\
&& - Bs \cos \phi - \beta s^2 \sin^2 \phi - \beta s \cos^2 \phi.
\end{array}
\label{eq:hamiltinian_s}
\end{equation}
%
We also assume macroscopically large number of particles, i.e. the $s = N/2 >> 1$ is C - number. Using (\ref{eq:hamiltinian_s})  for Schrodinger equation we obtain   
%
\begin{equation}
\begin{array}{l}
(\alpha - \beta \sin^2 \phi) \dfrac{d^2 \Phi}{d \phi^2} + (\beta s\sin{2\phi}-B\sin{\phi})) \dfrac{d \Phi}{d \phi} \\
+ (E + B s \cos \phi + \beta s^2 \sin^2 \phi) \Phi = 0,
\end{array}
\label{eq:schrodinger}
\end{equation}
%
where $\Phi \equiv \Phi(\phi)$ is unknown $2\pi$-periodic wavefunction that characterizes quantum phase properties, cf. \cite{Anglin}.
%ng that Eqs. (\ref{eq:hamiltinian_s}), (\ref{eq:schrodinger}) can be obtained by applying so-called phase-state representation, {\red cf. [  ]}.
%
%\begin{equation}
%\ket{\psi} = \dfrac{1}{2 \pi} \int_{-\infty}^{+\infty} d \phi \Phi(\phi) \ket{\phi}
%\end{equation}
%
%where
%
%\begin{equation}
%\ket{\phi} = \dfrac{1}{\sqrt{N!}} \Big[ e^{i\phi/2} \psi_1^\dagger + e^{-i \phi/2} \psi_2^\dag \Big]^N \ket{0}_1\ket{0}
%\end{equation}
%
%is macroscopic coherent spin-state (macroscopic qubit state) that plays important role in current quantum information protocols operating with $N$-particle condensates, cf. \cite{Byrnes_2012}.
It is possible to eliminate the term with first derivative in Eq. (\ref{eq:schrodinger}) applying the transformation
%
\begin{equation}
\Phi(\phi) = \Psi(z) \exp \Big( s \ln \dn{z} - \dfrac{\Lambda s}{2 \sqrt{\lambda (1 - \lambda)}} \arctan \Big( \sqrt{\dfrac{\lambda}{1 - \lambda}} \cn{z} \Big) \Big),
\label{eq:transformation_of_phi}
\end{equation}
%
where $z = \int \limits_0^\phi \dfrac{d \xi}{\sqrt{1 - \lambda \sin^2 \xi}} = F(\phi, \lambda)$ is new phase variable; $F(\phi, \lambda)$ is incomplete elliptic integral of the first kind.
In (\ref{eq:transformation_of_phi}) we have introduced dimension-less parameters $\Lambda = \dfrac{B}{\alpha s}$, $\lambda = \dfrac{\beta}{\alpha}$ playing crucial role for our model.
Inserting (\ref{eq:transformation_of_phi}) into (\ref{eq:schrodinger}) we arrive to Schr\"odinger-like equation
% 
\begin{equation}
\alpha \frac{d^2\Psi}{dz^2} + (E - V(z))\Psi = 0
\label{eq:schrodinger_usual}
\end{equation}
%
characterizing quantum particle with the ``mass''  $m = \dfrac{\hbar^2}{2 \alpha}$ and energy $E$, that moves in the potential $V(z) = \alpha s^2 V_0(z)$ with
%
\begin{equation}
V_0(z) = \frac{ (\frac{1}{4} \Lambda^2 - \lambda (1 - \lambda)) \sn^2{z} - \Lambda \cn{z}}{\dn^2{z}}.
\label{eq:potential}
\end{equation}
%

The dependences of trapping potential $V_0(z)$ on phase variable $z$ are shown in Fig. \ref{phase_potential}. The period of the functions is  $4K(\lambda)$.
The dependence of the $\lambda$ parameter on normalized half of inter-well distance $\dfrac{x_0}{a}$ can be estimated as $\lambda = 0.5 \Big( \exp \Big[ \dfrac{2 x_0^2}{a^2} \Big] - 1 \Big)^{-1}$. The results known for quantum phase mesoscopic Josephson junction model can be recovered from (\ref{eq:hamiltonian_spin}), (\ref{eq:potential}) at $\lambda = 0$ and shown by red curves in  Fig.\ref{phase_potential}. This limit corresponds to  infinitely large inter-well distances, with $x_0 \to \infty$.
On the other hand, $\lambda$ -- parameter grows rapidly at $\dfrac{x_0}{a} \ll 1$.
Obviously, in this limit our EPJJ model that based on assumption of relatively weak coupling between the wells breaks down.

Below we are focusing on $\lambda$ -- parameters which belong to segment of $0 < \lambda < 1$ and might be obtained at moderate values of $\dfrac{x_0}{a}$ which are $\dfrac{x_0}{a} \geq 0.45$.

%
\begin{figure}[ht]
\begin{center}\includegraphics[width=1\linewidth]{pic/potentials.eps}
\end{center}
\caption{
Effective dimension-less potential $V_0(z)$ for (a) $\Lambda = 100$, (b) $\Lambda = 0.5$ , and (c) $\Lambda = 0.01$ as a function of elliptic integral phase coordinate $z$.
Points $\pm z_{min}$ and $\pm z_{max}$ in (b) relay to global minima and maxima of the potential, correspondingly, occurring with periodicity $4K(\lambda)$. \label{phase_potential}}
\end{figure}
%
Analysis of the quantum phase can be performed  in three domains determined by vital $\Lambda$--parameter values, cf. \cite{Anglin}.

\textit{Rabi regime} $\Lambda \gg 1$. In this limit the trapping potential reads $V_R(z) = \dfrac{1}{4} \Lambda^2 \sd^2{z}$.
Physically, for any value of $\lambda$ -- parameter taken from the domain 0$<\lambda <1$ we can obtain (Rabi) oscillations for quantum phase.
In particular, small amplitude oscillations which are inherent to familiar Rabi regime can be achieved for negligable $\lambda$ --- see Fig. \ref{phase_potential}a and cf. \cite{Anglin}.

\textit{Fock regime} $\Lambda \ll 1/N^2$.
In purely quantum limit 
%the two mode system approaches by the state $\ket{\psi} \propto (\psi_1^\dag)^s (\psi_1^\dag)^s \ket{0}_1 \ket{0}_2$%
the phase wavefunction approaches $\Phi(\phi) \to (2\pi)^{-1/2}$.
In this regime, when inequality $\Lambda \ll \lambda < 1$ is holds, Eq.(\ref{eq:schrodinger_usual}) transforms into Mathieu equation that implies spectrum consisting from all Mathieu functions.

\textit{Josephson regime} $1/N^2\ll\Lambda <1$.
This regime represents intermediate case between Rabi and Fock limits --- see Fig. \ref{phase_potential}b.
The behavior of the phase depends on the ratio between parameters $\Lambda$ and $\lambda$.
If $\Lambda \ge 2\lambda$ the $V_0(z)$ posses only one minima at $z = 0$ and $V(0) = -\Lambda$. 
%Frequency of small amplitude ($z \ll 1$) oscillations %can be obtained from (\ref{eq:potential}) and looks like { \red $\omega_{0, quantum} = \frac{2 \alpha s}{\hbar} \sqrt{\frac{1}{4} \Lambda^2 - \frac{\Lambda}{2} (2\lambda - 1) - \lambda(1 - \lambda)}$ }.

In the other limit, for $\Lambda < 2 \lambda$, quantum "particle" is trapped at two minima of $V_0(z)$ that corresponds to $W$ -- like potential with coordinates $\cn{z_{min}} = \frac{\Lambda}{2 \lambda}$ and appearing for both of Josephson and Fock regimes, cf. Fig. \ref{phase_potential}b,c.  
The depth of the potential minima depends on the $\lambda$ parameter.
It is interesting to note that for $\Lambda > 2(1 - \lambda)$ potential $V_0(z)$ presumes local minimum at $\pm 2 K(\lambda)$, see Fig. 1b.

% $ce_0(z, q)$, $ce_1(z, q)$, $se_1(z, q)$, \dots, which are $2\pi$-periodic, and corresponding eigenvalues $a$ can be found using series expansions in powers of $q$.
%from (\ref{eq:potential}) one can obtain
%
%{\red

%
%\begin{equation}
%V_F(z) \simeq -\lambda (1 - \lambda) \sin^2{z},
%\end{equation}
%
%and our 
%transforms into Mathieu equation;
%
%\begin{equation}
%\frac{d^2 \Psi}{d z^2} + (a - 2q  \cos{2z}) \Psi = 0.
%\end{equation}
%
%where $a = E / \alpha + s^2 \lambda(1 - \lambda) / 2$, $q = s^2 \lambda (1 - \lambda)/ 4$.
%We recall our boundary condition: $\Psi(z + 2\pi) = \Psi(z)$.

%In another limit $\lambda \ll \Lambda$ 
%that might  be studied within Fock regime is relevant to condition . In this case from (\ref{eq:potential}) 
%we have $V_0(z) \simeq -\Lambda \cos{z}$.
%In fact, this limit that deals with negligible $\lambda$ reproduces results for convenient quantum phase model \cite{Anglin}.

@@

In current experiments with exciton polaritons  the condensate is observed at finite and high enough temperatures \cite{Sanvitto,Guillet}.
In this section we find conditions when quantum approach to the phase problem for Josephson junction with coupled exciton polariton condensates is valid. This conditions, in fact, responsible for ability of our system to perform quantum annealing algorithm \cite{Das}.

>>>>

\begin{abstract}
We consider finite temperature effects in a non-standard Bose-Hubbard model for an exciton polariton Josephson junction (EPJJ) that is characterised by complicated potential energy landscapes (PEL) consisting of sets of barriers and wells.
We show that the transition between  thermal  activation (classical) and tunneling (quantum) regimes exhibits universal features of the first and second order phase transition (PT) depending on the PEL for two polariton condensates.
In the presence of dissipation the relative phase of two condensates exhibits non-equilibrium PT from the quantum regime characterized by efficient tunneling of polaritons to the regime of permanent Josephson or Rabi oscillations, where the tunneling is suppressed, respectively.
This analysis paves way for the application of coupled polariton condensates for the realisation of a quantum annealing algorithm.

@@

In the XXI century, the studies of exciton-polariton Bose-Einstein condensates (BEC) in various type of semiconductor microstructures have become an important area of research in photonics and semiconductor physics, see e.g. \cite{Sanvitto, Guillet}.
Microcavity exciton polaritons are bosonic quasiparticles representing admixtures of quantized cavity photons and quantum well excitons.
Semiconductor microcavities are promising for various optoelectronic applications where the quantum  matter-field interface plays a crucial role.
In particular, it is worth to mention experimental demonstrations of polariton lasers with electrical injection \cite{Bhattacharya,Schneider}, polariton amplifiers \cite{Niemietz}, switches \cite{Amo_2010}, transistors \cite{Ballarini}, polariton circuits and optical logic elements \cite{Sturm,Liew}.
Although exciton polariton condensates exhibit the Bose-Einstein statistics above the lasing threshold and are characterized by a macroscopic occupation of the ground state at certain pumping rate that is less than the threshold pumping for convenient lasers, they are not in a true thermodynamic equilibrium state \cite{Byrnes_2014,Sun}. Non-equilibrium features of exciton polariton condensates play an important role in various manifestations of their collective (many body) states such as superfluidity \cite{Carusotto_2013,Amo_2009}, quantized vortices \cite{Lagoudakis_2008,Lagoudakis_2009}, soliton formation \cite{Sich}, Josephson oscillations and macroscopic self-trapping \cite{Abbarchi,Lagoudakis_2010}.

The problem of distinguishability of statistically classical (thermal) and quantum regimes for exciton polariton condensates is very important for their possible applications in quantum information technologies, see e.g. \cite{Dominici}.
Fast switching properties (the typical switching time of a few picoseconds), relatively strong nonlinear response and flexibility to external optical and/or electrical pumping spin degrees of freedom made microcavity polaritons potentially promising for quantum computation and quantum information processing, see, e.g. \cite{Demirchyan,Pagel,Kyriienko,Solnyshkov_2015, Dominici}.

Here we specifically focus on the quantum annealing problem that is relevant to the searching algorithm for the global minimum of the potential energy landscape (PEL) consisting of a set of barriers and wells, see e.g.  \cite{Santoro, Das}.
In a purely classical  (thermal) regime the {bosonic quasipartcles cross barriers stochastically} at finite temperature with the help of thermal activation process if the thermal energy is large enough, see Fig. \ref{pic:potential_picture}. Contrary, in a quantum limit the same system undergoes quantum tunneling through the barrier. Obviously, the shape (height and thickness) of the barrier plays an essential role in this case \cite{Das, Lewenstein}. It has been shown recently in \cite{Lewenstein} that the annealing algorithm relies in general case on the combination of thermal annealing and quantum tunneling. The     
collective (bosonic) character of polariton condensation could be used for the acceleration of the physical implementation of the search algorithm \cite{Yan}. The authors of Refs. \cite{Lago1, Lago2} propose polariton graphs as an analog platform for minimizing the XY -- Hamiltonian by exploring the non-equilibrium character of exciton polariton condensates and mimicking various magnetic phases with them.  An original technique of nonresonant spatially  modulated pumping beam  have been used to imprint two-dimensional polariton graphs with different topology and different heights of potential barriers. Importantly, the phase-locking  between arbitrary neighboring vertices might be practically achieved in this case.
%
\begin{figure}[ht]
\center{\includegraphics[width=0.5\linewidth]{pic/potential_picture.eps}}
\caption{Scketch of two tunnel coupled exciton condensates. The two minima of the trapping potential $U$ are situated in the points $+x_0$ and $-x_0$. At thermal equilibrium the interplay between thermal and quantum annealing effects is governed by the blue shift of condensates, effective temperature of exciton-polaritons and the shape of the potential. \label{pic:potential_picture}}
\end{figure} 
%
In the present paper we study the effect of a finite effective temperature on the coupling of quasi-equilibrium exciton-polariton condensates accounting for the competing thremal and quantum annealing effects.

Recently Yongbao Sun et al \cite{Snoke_2017} reported the observation of a quasi-equilibrium low branch exciton polariton condensate in a high Q-factor microcavity characterised by a remarkably long photon lifetime of $135ps$.
Clearly, exciton polaritons with long lifetime are promising candidates for quantum technologies \cite{Demirchyan}.
One of the crucial advantages of exciton-polaritons over cold atoms consists in a perspective of room temperature operation of polariton devices.
In this context, it is very important to reveal the impact of temperature on the physics of interacting polariton condensates.
It should be also noted that the effective temperature of a polariton gas may be introduced in the quasi-equilibrium approximation, which is not necessarily valid in all cases but constitutes an important starting point of any analysis. Further steps would involve a full kinetic modelling for a non-equilibrium polariton gas.

To be specific, here we consider an exciton polariton Josephson junction (EPJJ).
We study the interplay of two mechanisms of coupling between the condensates in order to reveal the cross-over from the quantum tunneling to the incoherent coupling regime that may occur at finite temperatures \cite{Aleiner, Shelykh_2008, Borgh_2010}. 
In particular, we are interested in a generic problem of the quantum-classical phase transition (PT) applied to two-level (spin) systems \cite{Chudnovsky_1997, Caldeira, Larkin, Riseborough, Zhang, Owerre}.
Traditionally,  studies in this field are limited by consideration of superconductor devices \cite{Ankerhold} where macroscopic quantum tunneling phenomena plays an essential role.
In particular, it is worth to mention here the Schr\"odinger cat state formation \cite{Leggett} and the 	design of quantum gates with superconductor qubits for quantum computing where a macroscopic quantum coherence for the phase is important \cite{Makhlin}.

We shall consider a generalised model of EPJJ that relates to so-called non-standard Bose-Hubbard models (see \cite{Dutta} and references therein) where the energy of polariton-polariton scattering contributes to the tunneling parameter matrix elements \cite{Aleiner, Shelykh_2008, Borgh_2010, Solnyshkov_2008, Sarchi}.
In particularly, we pay attention to the temperature dependent quantum critical phenomena occurring in the presence of macroscopic tunneling.
We aim at the formulating of the criteria for realisation of the quantum tunneling regime that is crucial for the realisation of the quantum annealing algorithms and exciton-polariton quantum gates.

This paper is organised as follows.
In Sec. \ref{sec:quantum_phase} we introduce the quantum phase model for EPJJ and discuss various regimes of coupling between two condensates at zero temperature and in the limit of an infinite exciton polariton lifetime.
Section \ref{sec:quantum_classical} presents the results obtained for a quantum-classical PT problem and addresses the quantum phase at finite temperatures.
The influence of non-equilibrium effects involving the microcavity exciton polaritons possessing a finite lifetime is discussed in Sec. \ref{sec:non-equilibrium}.      
In Conclusion, we summarize the obtained results.
In the Supplementary Material we explain in details the extended model of EPJJ formulated within the tight-binding approximation.

@@

We consider a system of two spinless exciton-polariton condensates localised in lateral potential traps created in a planar semiconductor microcavity.
The condensates are confined by a W-shape potential $U(x)$ possessing two minima at the points $\pm x_0$, see Fig.1.
Polaritons are pumped to the condensates from a reservoir of incoherent excitons. In this and next Section, we shall neglect the finite polariton lifetime and assume that the polariton gas is at the thermal equilibrium.   
We account for the polariton-polariton repulsion.

We shall use the pseudo-spin operator representation for the Hamiltonian describing our EPJJ model system that has a form (the details of the model are presented in the Supplementary Material): 
%
\begin{equation}
\hat{H} = \alpha \hat{S}_z^2 + \beta \hat{S}_x^2 - B \hat{S}_x.
\label{eq:hamiltonian_spin}
\end{equation}
%
The coefficients in Eq. (1) characterize the tunnel coupling of two trapped condensates:
the parameters $\alpha$ and $\beta$ are proportional to the overlap integrals of real symmetric ($\Phi_+$) and antisimmetric ($\Phi_-$) wavefunctions of two condensates that are
$\int \Phi_i^2 \Phi_j^2 dx, i,j \in \{+,-\}$, $B$ is determined by difference of chemical potentials which obey stationary Gross-Pitaevskii (GP) equations for two trapped condensates.

It is important to note that the last term in Eq. (1) characterizes the familiar XY--model Hamiltonian  considered in
\cite{Lago1, Lago2}.
The second term in (1) introduces an additional part that has no analogy in the XY-Hamiltonian; it is proportional to $\cos(2\phi)$, where $\phi$ is the phase difference between two polariton condensates.
We will show that this term becomes essential if the barrier separating two condensates is sufficiently large.

It is convenient to introduce the pseudo-spin operators in a form
% 
\begin{subequations}
\begin{align}
\hat{S}_x = & s \cos{\phi} - \sin{\phi} \dfrac{d}{d \phi}, \\
\quad \hat{S}_y = & s \sin{\phi} + \cos{\phi} \dfrac{d}{d \phi}, \\
\hat{S}_z = & -i\dfrac{d}{d \phi},
\end{align}
\label{eq:pseudo_spin_diff}
\end{subequations}
%
The operators defined in Eq. (\ref{eq:pseudo_spin_diff}) obey familiar $SU(2)$ algebra commutation relations $[\hat{S}_i, \hat{S}_j] = i\epsilon_{ijk}\hat{S}_k,  i,j,k=x,y,z$.
After some straightforward calculations from (\ref{eq:hamiltonian_spin}) one can obtain
%
\begin{equation}
\begin{array}{lcl}
H & = & -(\alpha - \beta \sin^2 \phi) \dfrac{d^2}{d \phi^2} - (\beta(s-\dfrac{1}{2})\sin{2\phi} - B\sin{\phi}) \dfrac{d}{d \phi} \\
&& - Bs \cos \phi - \beta s^2 \sin^2 \phi - \beta s \cos^2 \phi.
\end{array}
\label{eq:hamiltinian_s}
\end{equation}
%
Thereafter we assume that both condensates are composed by macroscopically large numbers of particles, i.e. the $s = N/2 >> 1$ is C - number.
The general stationary Schroedinger equation with a Hamiltonian (\ref{eq:hamiltinian_s}) writes:
%
\begin{equation}
\begin{array}{l}
(\alpha - \beta \sin^2 \phi) \dfrac{d^2 \Phi}{d \phi^2} + (\beta s\sin{2\phi}-B\sin{\phi})) \dfrac{d \Phi}{d \phi} \\
+ (E + B s \cos \phi + \beta s^2 \sin^2 \phi) \Phi = 0,
\end{array}
\label{eq:schrodinger}
\end{equation}
%
where $\Phi \equiv \Phi(\phi)$ is some $2\pi$-periodic wavefunction that characterizes quantum phase properties \cite{Anglin}.
It is possible to eliminate the term with first derivative in Eq. (\ref{eq:schrodinger}).
In order to find the explicite form of the function $\Phi$, it is convenient to substitute it by:
%
\begin{equation}
\Phi(\phi) = \Psi(z) \exp \Big( s \ln \dn{z} - \dfrac{\Lambda s}{2 \sqrt{\lambda (1 - \lambda)}} \arctan \Big( \sqrt{\dfrac{\lambda}{1 - \lambda}} \cn{z} \Big) \Big),
\label{eq:transformation_of_phi}
\end{equation}
%
where $z = \int \limits_0^\phi \dfrac{d \xi}{\sqrt{1 - \lambda \sin^2 \xi}} = F(\phi, \lambda)$ is a new phase variable; $F(\phi, \lambda)$ is the incomplete elliptic integral of the first kind.
In (\ref{eq:transformation_of_phi}) we have introduced dimension-less parameters $\Lambda = \dfrac{B}{\alpha s}$, $\lambda = \dfrac{\beta}{\alpha}$.
Inserting (\ref{eq:transformation_of_phi}) into (\ref{eq:schrodinger}) we arrive to the familiar form of a Schr\"odinger equation
% 
\begin{equation}
\alpha \frac{d^2\Psi}{dz^2} + (E - V(z))\Psi = 0.
\label{eq:schrodinger_usual}
\end{equation}
%
This equation describes an effective particle with the mass  $m = \dfrac{\hbar^2}{2 \alpha}$ and energy $E$, that is confined by the potential $V(z) = \alpha s^2 V_0(z)$ with
%
\begin{equation}
V_0(z) = \frac{ (\frac{1}{4} \Lambda^2 - \lambda (1 - \lambda)) \sn^2{z} - \Lambda \cn{z}}{\dn^2{z}}.
\label{eq:potential}
\end{equation}
%

The dependences of the trapping potential $V_0(z)$ on the phase variable $z$ are shown in Fig. \ref{phase_potential}. The period of the functions is  $4K(\lambda)$.
The dependence of the $\lambda$ parameter on normalized half of the inter-well distance $\dfrac{x_0}{a}$ can be approximated by $\lambda = 0.5 \Big( \exp \Big[ \dfrac{2 x_0^2}{a^2} \Big] - 1 \Big)^{-1}$.
The well-known behaviour of the quantum phase mesoscopic Josephson junction (that map to the XY--model, \citep{Lago1, Lago2} ) can be recovered from (\ref{eq:hamiltonian_spin}), (\ref{eq:potential}) at $\lambda = 0$ and shown by red curves in  Fig.\ref{phase_potential}. This limit corresponds to  infinitely large inter-well distances, with $x_0 \to \infty$.
On the other hand, the $\lambda$ -- parameter exhibits a sharp increase at $\dfrac{x_0}{a} \ll 1$.
Obviously, in this limit our EPJJ model based on the assumption of a relatively weak coupling between trapped condensates breaks down.
Below we are focusing on $\lambda$ -- parameters which belong to the range of $0 < \lambda < 1$ and correspond to the moderate values of $\dfrac{x_0}{a}$ such as $\dfrac{x_0}{a} \geq 0.45$. 
%
\begin{figure}[ht]
\begin{center}\includegraphics[width=1\linewidth]{pic/potentials.eps}
\end{center}
\caption{
Effective dimension-less potential $V_0(z)$ for (a) $\Lambda = 100$, (b) $\Lambda = 0.5$ , and (c) $\Lambda = 0.01$ as a function of the elliptic integral phase coordinate $z$.
Points $\pm z_{min}$ and $\pm z_{max}$ in (b) correspond to global minima and maxima of the potential, respectively. \label{phase_potential}}
\end{figure}
%
The analysis of the quantum phase can be performed in three domains determined by vital $\Lambda$--parameter values, \cite{Anglin}.

\textit{Rabi regime} $\Lambda \gg 1$. In this limit the trapping potential can be approximated by $V_R(z) = \dfrac{1}{4} \Lambda^2 \sd^2{z}$.
Physically, for any value of $\lambda$ -- parameter belonging to the domain 0$<\lambda <1$ the system exhibits (Rabi) oscillations for the phase.
In particular, small amplitude oscillations which are inherent to familiar Rabi regime can be achieved for negligable $\lambda$ --- see Fig. \ref{phase_potential}a and \cite{Anglin}.

\textit{Fock regime} $\Lambda \ll 1/N^2$.
In purely quantum limit $\Phi(\phi) \to (2\pi)^{-1/2}$.
In this regime, when inequality $\Lambda \ll \lambda < 1$ is holds, Eq.(\ref{eq:schrodinger_usual}) transforms to the Mathieu equation that is characterised by the set of eignen states consisting of all Mathieu functions.

\textit{Josephson regime} $1/N^2\ll\Lambda <1$.
This regime corresponds to the intermediate case between Rabi and Fock limits --- see Fig. \ref{phase_potential}b.
The behavior of the phase depends on the ratio between the parameters $\Lambda$ and $\lambda$.
If $\Lambda \ge 2\lambda$ the potential $V_0(z)$ possesses only one minimum at $z = 0$ and $V(0) = -\Lambda$.

In the opposite limit, for $\Lambda < 2 \lambda$, the effective quantum ``particle'' described by Eq. (\ref{eq:schrodinger_usual}) is trapped at the two minima of $V_0(z)$ that correspond to the $W$ -- like potential with coordinates $z_{min} = \pm \cn^{-1}{\frac{\Lambda}{2 \lambda}}$ and appearing for both of Josephson and Fock regimes, see Fig. \ref{phase_potential}b,c.  
The depth of the potential minima depends on the $\lambda$ parameter.
It is interesting to note that for $\Lambda > 2(1 - \lambda)$ the potential $V_0(z)$ possesses a local minimum at $\pm 2 K(\lambda)$, see Fig. 1b.

@@

Experimentally, exciton-polariton condensates are being observed at elevated temperatures, un to the room temperature in wide-gap semiconductor microcavities \cite{Sanvitto,Guillet}.
At high temperatures the validity of the quantum coherent treatment of the Josephson problem is limited.
In this section, we study the limits of validity of the quantum approach which also set the limits of the system suitability for quantum annealing \cite{Das}.
